Weighted Model Counting with Conditional Weights for Bayesian Networks

### 2020-08-27 paper/paper.tex
Some editing

\def\year{2021}\relax
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage{aaai21}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet} % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{graphicx}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in}  % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in}  % DO NOT CHANGE THIS
%\nocopyright
%PDF Info Is REQUIRED.
% For /Author, add all authors within the parentheses, separated by commas. No accents or commands.
% For /Title, add Title in Mixed Case. No accents or commands. Retain the parentheses.
\pdfinfo{
/Title (Weighted Model Counting with Conditional Weights)
/Author ()
/TemplateVersion (2021.1)
} %Leave this
% /Title ()
% Put your actual complete title (no codes, scripts, shortcuts, or LaTeX commands) within the parentheses in mixed case
% Leave the space between \Title and the beginning parenthesis alone
% /Author ()
% Put your actual complete list of authors (no codes, scripts, shortcuts, or LaTeX commands) within the parentheses in mixed case.
% Each author should be only by a comma. If the name contains accents, remove them. If there are any LaTeX commands,
% remove them.

@@

\title{Weighted Model Counting with Conditional Weights}
%\author{
%  Paulius Dilkas, \textsuperscript{\rm 1}
%  Vaishak Belle \textsuperscript{\rm 1,2} \\
%}
%\affiliations{
%    \textsuperscript{\rm 1} University of Edinburgh \\
%    \textsuperscript{\rm 2} Alan Turing Institute \\
%    p.dilkas@sms.ed.ac.uk, vaishak@ed.ac.uk
%}
\begin{document}

@@

The most well-studied version of WMC assigns weights to models based on
weights on literals, i.e., the weight of a model is the product of the weights
of all literals in it. This simplification is motivated by the fact that the
number of models scales exponentially with the number of atoms, so listing the
weight of every model is intractable. However, this also severely restricts what
probability distributions can be represented. A common way to overcome this
limitation is by adding more literals. While we show that this is always
possible in \cref{sec:wmc_as_measure}, the rest of the paper demonstrates that
it can be significantly more efficient to encode weights in a more flexible
format instead.

\Cref{sec:prelims} describes three equivalent perspectives on the subject based
on logic, set theory, and Boolean algebras. Furthermore, we describe the space
of functions on Boolean algebras and various operations on those functions.
\Cref{sec:wmc_as_measure} introduces WMC as the problem of computing the value
of a measure on a Boolean algebra. We show that not all measures can be
represented using literal-based WMC, but all Boolean algebras can be extended to
make any measure representable in such a manner.

In \cref{sec:bns}, we introduce \emph{conditional weights} on literals (in the
spirit of conditional probabilities) that have literal-based WMC as a special
case. As a particular example, we show how conditional weights can be used to
encode Bayesian networks. We prove the correctness of the encoding and show
how a state-of-the-art WMC solver ADDMC \cite{DBLP:conf/aaai/DudekPV20} can be
adapted to the new format. ADDMC is a recently-proposed algorithm for WMC based
on manipulating functions on Boolean algebras using an efficient representation
for such functions known as \emph{algebraic decision diagrams} (ADDs)
\cite{DBLP:journals/fmsd/BaharFGHMPS97}.

Finally, in \cref{sec:experiments}, we show how the new encoding results in
significantly improved inference speed on most benchmark instances, often by one
or two orders of magnitude. We explain the performance benefits by showing how
the new encoding has asymptotically fewer variables and ADDs.

@@

In this section, we give a brief introduction to two alternative ways to think
about logical constructs such as models and formulas. Let $\mathcal{L}$ be a
propositional logic with atoms $a$ and $b$, and let $U = \{ a, b \}$. Then
$2^U$, the power set of $U$, is the set of all models of $\mathcal{L}$, and
$2^{2^U}$ is the set of all formulas. These sets can also be represented as
Boolean algebras (e.g., using the syntax $(2^{2^U}, \land, \lor, \neg, \bot,
\top)$) with a partial order $\le$ that corresponds to set inclusion
$\subseteq$---see \cref{tbl:notation_example} for examples of how various
elements can be represented in both notations. Most importantly, note that the
word \emph{atom} has completely different meanings in logic and in Boolean
algebras. An atom in $\mathcal{L}$ is an atomic formula, i.e., an element of
$U$, whereas an atom in a Boolean algebra is (in set-theoretic terms) a
singleton set. For instance, an atom in $2^{2^U}$ corresponds to a model of
$\mathcal{L}$, i.e., an element of $2^U$. Unless referring specifically to a
logic, we will use the algebraic definition of an atom and refer to logical
atoms as \emph{variables}. In the rest of the paper, for any set $U$, we will
use set-theoretic notation for $2^U$ and Boolean-algebraic notation for
$2^{2^U}$, except for (Boolean) atoms in $2^{2^U}$ that are denoted as $\{x\}$
for some model $x \in 2^U$.

@@

\begin{observation}
  Let $U$ be a set, and $\mathcal{V} = \{ \alpha\colon 2^X \to \mathbb{R}_{\ge 0}
  \mid X \subseteq U \}$. Then $\mathcal{V}$ is a semi-vector space with three
  additional operations: (non-scalar) multiplication, complement, and projection.
  Specifically, note that both addition and multiplication are both associative
  and commutative.
\end{observation}

@@

\Cref{alg:encoding} shows how a Bayesian network with vertices $\mathcal{V}$
can be represented as a weight function $\phi\colon 2^U \to \mathbb{R}_{\ge 0}$.
The algorithm begins with the unit function and multiplies it by
$\mathrm{CPT}_X\colon 2^{\mathcal{E}^*(X)} \to \mathbb{R}_{\ge 0}$ for each
random variable $X \in \mathcal{V}$. We call each such function a
\emph{conditional weight function} as it represents a conditional probability
distribution. However, the distinction is primarily a semantic one: a function
$2^{\{a, b\}} \to \mathbb{R}_{\ge 0}$ can represent $\Pr(a \mid b)$, $\Pr(b \mid
a)$, or something else entirely.

@@

In this section, we describe an experimental study comparing the five WMC
encodings for Bayesian networks when run with ADDMC. The experiments were run on
servers with Intel Xeon Gold 6138 and Intel Xeon E5-2630
processors\footnote{Each instance is run on the same processor for all
  encodings.} with an \SI{8}{\gibi\byte} memory limit and a \SI{1000}{\second}
timeout.

@@

For experimental data, we use the Bayesian networks available with Ace and
Cachet\footnote{\url{https://www.cs.rochester.edu/u/kautz/Cachet/}}, most of
which happen to be binary. We classify them into the following seven categories:
\begin{itemize*}
\item DQMR and
\item Grid networks as described by \citet{DBLP:conf/aaai/SangBK05},
\item Mastermind, and
\item Random Blocks from the work of \citet{DBLP:journals/ijar/ChaviraDJ06},
\item remaining binary Bayesian networks that include Plan Recognition
  \cite{DBLP:conf/aaai/SangBK05}, Friends and Smokers, Students and Professors
  \cite{DBLP:journals/ijar/ChaviraDJ06}, and \texttt{tcc4f}, and
\item non-binary classic Bayesian networks (\texttt{alarm}, \texttt{diabetes},
  \texttt{hailfinder}, \texttt{mildew}, \texttt{munin1}--\texttt{4},
  \texttt{pathfinder}, \texttt{pigs}, \texttt{water}).
\end{itemize*}

@@

While we chose to use ADDMC \cite{DBLP:conf/aaai/DudekPV20} as the WMC algorithm
and Bayesian networks as a canonical example of a probabilistic inference task,
these are only examples meant to illustrate the broader idea that choosing a
more expressive representation of weights can outperform increasing the size of
the problem to keep the weights simple. The same idea can be adapted to
search-based solvers such as Cachet \cite{DBLP:conf/sat/SangBBKP04} as well as
to other inference problem domains such as probabilistic programs.

@@

\appendix
\section{Proofs}

@@

\bibliography{paper}
\end{document}

>>>>

\def\year{2021}\relax
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage{aaai21}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet} % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{graphicx}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in}  % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in}  % DO NOT CHANGE THIS
%\nocopyright
%PDF Info Is REQUIRED.
% For /Author, add all authors within the parentheses, separated by commas. No accents or commands.
% For /Title, add Title in Mixed Case. No accents or commands. Retain the parentheses.
\pdfinfo{
/Title (Weighted Model Counting with Conditional Weights for Bayesian Networks)
/Author ()
/TemplateVersion (2021.1)
} %Leave this
% /Title ()
% Put your actual complete title (no codes, scripts, shortcuts, or LaTeX commands) within the parentheses in mixed case
% Leave the space between \Title and the beginning parenthesis alone
% /Author ()
% Put your actual complete list of authors (no codes, scripts, shortcuts, or LaTeX commands) within the parentheses in mixed case.
% Each author should be only by a comma. If the name contains accents, remove them. If there are any LaTeX commands,
% remove them.

@@

\title{Weighted Model Counting with Conditional Weights for Bayesian Networks}
%\author{
%  Paulius Dilkas, \textsuperscript{\rm 1}
%  Vaishak Belle \textsuperscript{\rm 1,2} \\
%}
%\affiliations{
%    \textsuperscript{\rm 1} University of Edinburgh \\
%    \textsuperscript{\rm 2} Alan Turing Institute \\
%    p.dilkas@sms.ed.ac.uk, vaishak@ed.ac.uk
%}
\begin{document}

@@

The most well-studied version of WMC assigns weights to models based on
weights on literals, i.e., the weight of a model is the product of the weights
of all literals in it. This simplification is motivated by the fact that the
number of models scales exponentially with the number of atoms, so listing the
weight of every model is intractable. However, this also severely restricts what
probability distributions can be represented. A common way to overcome this
limitation is by adding more literals. While we show that this is always
possible, we also demonstrate that it can be significantly more efficient to
encode weights in a more flexible format instead.

After briefly reviewing the background in \cref{sec:related}, in
\cref{sec:prelims} we  describe three equivalent perspectives on the subject
based on logic, set theory, and Boolean algebras. Furthermore, we describe the
space of functions on Boolean algebras and various operations on those
functions. \Cref{sec:wmc_as_measure} introduces WMC as the problem of computing
the value of a measure on a Boolean algebra. We show that not all measures can
be represented using literal-based WMC, but all Boolean algebras can be extended
to make any measure representable in such a manner.

This new measure-theoretic perspective allows us to not only encode any
discrete probability distribution but also improve inference speed. In
\cref{sec:bns}, we demonstrate this by developing a new WMC encoding for
Bayesian networks that uses \emph{conditional weights} on literals (in the
spirit of conditional probabilities) that have literal-based WMC as a special
case. We prove the correctness of the encoding and show how a state-of-the-art
WMC solver ADDMC \cite{DBLP:conf/aaai/DudekPV20} can be adapted to the new
format. ADDMC is a recently-proposed algorithm for WMC based on manipulating
functions on Boolean algebras using an efficient representation for such
functions known as algebraic decision diagrams (ADDs)
\cite{DBLP:journals/fmsd/BaharFGHMPS97}. Finally, in \cref{sec:experiments}, we
show how the new encoding results in faster inference on the vast majority of
benchmark instances, often by one or two orders of magnitude. We explain the
performance benefits by showing how our encoding has asymptotically fewer
variables and ADDs.

@@

In this section, we give a brief introduction to two alternative ways to think
about logical constructs such as models and formulas. Let us consider a simple
example of a propositional logic $\mathcal{L}$ with only two atoms $a$ and $b$,
and let $U = \{ a, b \}$. Then $2^U$, the power set of $U$, is the set of all
models of $\mathcal{L}$, and $2^{2^U}$ is the set of all formulas. These sets
can also be represented as Boolean algebras (e.g., using the syntax $(2^{2^U},
\land, \lor, \neg, \bot, \top)$) with a partial order $\le$ that corresponds to
set inclusion $\subseteq$---see \cref{tbl:notation_example} for examples of how
various elements can be represented in both notations. Most importantly, note
that the word \emph{atom} has completely different meanings in logic and in
Boolean algebras. An atom in $\mathcal{L}$ is an atomic formula, i.e., an
element of $U$, whereas an atom in a Boolean algebra is (in set-theoretic terms)
a singleton set. For instance, an atom in $2^{2^U}$ corresponds to a model of
$\mathcal{L}$, i.e., an element of $2^U$. Unless referring specifically to a
logic, we will use the algebraic definition of an atom and refer to logical
atoms as \emph{variables}. In the rest of the paper, for any set $U$, we will
use set-theoretic notation for $2^U$ and Boolean-algebraic notation for
$2^{2^U}$, except for (Boolean) atoms in $2^{2^U}$ that are denoted as $\{x\}$
for some model $x \in 2^U$.

@@

Note that if $U$ is any set, then $\mathcal{V} = \{ \alpha\colon 2^X \to
\mathbb{R}_{\ge 0} \mid X \subseteq U \}$ is a semi-vector space with three
additional operations: (non-scalar) multiplication, complement, and projection.
Specifically, note that both addition and multiplication are both associative
and commutative.

@@

\Cref{alg:encoding} shows how a Bayesian network with vertices $\mathcal{V}$
can be represented as a weight function $\phi\colon 2^U \to \mathbb{R}_{\ge 0}$.
The algorithm begins with the unit function and multiplies it by
$\mathrm{CPT}_X\colon 2^{\mathcal{E}^*(X)} \to \mathbb{R}_{\ge 0}$ for each
random variable $X \in \mathcal{V}$. We call each such function a
\emph{conditional weight function} as it represents a conditional probability
distribution. However, the distinction is primarily a semantic one: a function
$2^{\{a, b\}} \to \mathbb{R}_{\ge 0}$ can represent $\Pr(a \mid b)$, $\Pr(b \mid
a)$, or something else entirely, e.g., $\Pr(a \land b)$, $\Pr(a \lor b)$, etc.

@@

In this section, we describe an experimental study comparing the five WMC
encodings for Bayesian networks when run with ADDMC. The experiments were run on
servers with Intel Xeon Gold 6138 and Intel Xeon E5-2630
processors\footnote{Each instance is run on the same processor for all
  encodings.} running Scientific~Linux~7 with an \SI{8}{\gibi\byte} memory limit
and a \SI{1000}{\second} timeout.

@@

For experimental data, we use the Bayesian networks available with Ace and
Cachet\footnote{\url{https://www.cs.rochester.edu/u/kautz/Cachet/}}, most of
which happen to be binary. We classify them into the following seven categories:
\begin{itemize*}
\item DQMR and
\item Grid networks as described by \citet{DBLP:conf/aaai/SangBK05},
\item Mastermind, and
\item Random Blocks from the work of \citet{DBLP:journals/ijar/ChaviraDJ06},
\item remaining binary Bayesian networks that include Plan Recognition
  \cite{DBLP:conf/aaai/SangBK05}, Friends and Smokers, Students and Professors
  \cite{DBLP:journals/ijar/ChaviraDJ06}, and \texttt{tcc4f}, and
\item non-binary classic Bayesian networks (\texttt{alarm}, \texttt{diabetes},
  \texttt{hailfinder}, \texttt{mildew}, \texttt{munin1}--\texttt{4},
  \texttt{pathfinder}, \texttt{pigs}, \texttt{water}).
\end{itemize*}
We run ADDMC with each of the five encodings once on each Bayesian network.

@@

While we chose to use ADDMC \cite{DBLP:conf/aaai/DudekPV20} as the WMC algorithm
and Bayesian networks as a canonical example of a probabilistic inference task,
these are only examples meant to illustrate the broader idea that choosing a
more expressive representation of weights can outperform increasing the size of
the problem to keep the weights simple. Indeed, in this work, we have provided a
new theoretical perspective on the expressive power of WMC and illustrated the
empirical benefits of that perspective. The same idea can be adapted to
search-based solvers such as Cachet \cite{DBLP:conf/sat/SangBBKP04} as well as
to other inference problem domains such as probabilistic programs
\cite{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:journals/corr/abs-2005-09089}.

@@

\newpage
\bibliography{paper}

\newpage
\appendix
\section{Proofs}

@@

\end{document}
